First Order Alternation

### 2018-06-29 quantifiers/draft.tex
ok

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of \emph{positive} first-order formulae is
defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \Rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

The execution semantics of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each
tree $T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\F$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}

\comment[ri]{I am here}

In several of the upcoming developments it is more convenient to work
with a symbolic execution semantics, that describes a set of execution
forests by a sequence of formulae obtained one from another using
rewriting rules. Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an
automaton and $\stamp{X}{i} = \set{\stamp{x}{i} \mid x \in X}$ be a
set of \emph{time-stamped} variables. For a formula $\psi$ and $i \in
\nat$, we define $\stamp{\psi}{i} = \psi[X/\stamp{X}{i}]$, i.e.\ the
formula in which each variable $x \in X$, denoting an input data
element, is replaced by $\stamp{x}{i}$. Further, given a formula
$\phi$, let $\Accept{F}{\phi}$ be the formula obtained from $\phi$ by
replacing each predicate atom $q(z_1,\ldots,z_{\#(q)})$ by $\top$ if
$q \in F$, and by $\bot$, otherwise.

For a sequence of input events $a_1 \ldots a_n \in \Sigma^*$, we
consider the sequence of formulae \(\phi_0 \stackrel{a_1}{\Rightarrow}
\phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
\phi_n\), where $\phi_0 = \iota$ and, for each $i \in [n-1]$, the
formula $\phi_{i+1}$ is obtained from $\phi_i$ by replacing each
predicate atom $q(t_1,\ldots,t_{\#(q)})$ from $\phi_i$ by the formula
$\stamp{\psi}{i+1}[t_1, \ldots, t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
the only transition rule that applies, or by $\bot$, if no transition
rule applies. The following lemma formalizes the relation between the
symbolic and the configuration-based semantics of automata:

\begin{lemma}\label{lemma:symbolic-execution}
  A word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is accepted by $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ if and only if $\I,\nu \models
  \Accept{F}{\phi_n}$ for each interpretation $\I$, where the
  valuation $\nu : \bigcup_{i=1}^n \stamp{X}{i} \rightarrow \Data$ is
  defined as $\nu(\stamp{x}{i}) = \nu_i(x)$, for each $x \in X$ and $i
  \in [n]$.
\end{lemma}
\proof{ ``$\Rightarrow$'' Assume that $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ is accepted by $\A$ and let $\F_w$ be an execution
  forest of $\A$ over $w$. First, it is easy to check that $\phi_i$ is
  a positive formula, for each $i \in [n]$. Next, we define the
  interpretation: \[\I^w_n(q) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level of }
    F_w}\] and prove $\I^w_n,\nu \models \phi_n$, by induction on
  $n\geq0$. For $n=0$, we have $\cube{\I^w_0} \in
  \cube{\sem{\iota}}$ and $\I^w_0 \models \iota$ follows, by
  definition. For the induction step, assume that $\I^w_{n-1},\mu
  \models \phi_{n-1}$, where $\mu(\stamp{x}{i}) = \nu_i(x)$, for each
  $x \in X$ and $i \in [n-1]$.

  \noindent''$\Leftarrow$'' 
\qed}

>>>>

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of first-order formulae is defined by the syntax
below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \Rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$. Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n) \in \Sigma[X]^*$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
such that:
\begin{compactitem}
\item the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\F$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{\A}{w_\Sigma}$. We show that there
  exists an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{\A}{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{\A}{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. We extend this to an
  execution over $w$, by appending to each configuration
  $q(d_1,\ldots,d_{\#(q)})$ on the frontier [($n-1$)-th level] of a
  tree $T \in \G$ the leaves corresponding to \qed}
